---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# PSLassoSynthNC

<!-- badges: start -->
<!-- badges: end -->

# Background

When fitting Lasso models for propensity score weighting, both theory and simulations have shown that undersmoothing the Lasso model is beneficial for reducing bias in estimated treatment effects. However, determining the degree of undersmoothing can be challenging. Proper undersmoothing can reduce bias in estimated effects, but excessive undersmoothing can increase the risk of bias caused by severe overfitting. This package uses balance diagnostics to determine the degree of undresmoothing when fitting Lasso models for propensity score weighting. However, tuning Lasso models using balance metrics are not guaranteed to minimize bias in PS weighted estimators since such metrics are outcome blind. Therefore, this package further provides functions to generate synthetic negative control exposures for bias detection. 

This package is based on work in the following paper:

* Wyss R, Hansen BB, Hahn G, van der Laan L, Lin KJ. Undersmoothed LASSO models for propensity score weighting and synthetic negative control exposures for bias detection. _arXiv:2506.17760[stat.ME]_. 2025.

Additional references on undersmoothing propensity score models are provided in the following:

* Ju C, Wyss R, Franklin JM, et al. Collaborative-controlled LASSO for constructing propensity score-based estimators in high-dimensional data. _Statistical Methods in Medical Research_. 2019;28(4):1044-1063.

* Ertefaie A, Hejazi NS, van der Laan MJ. Nonparametric inverse-probability-weighted estimators based on the highly adaptive lasso. _Biometrics_. 2023;79(2):1029-1041.

* Wyss R, van der Laan M, Gruber S, et al. Targeted learning with an undersmoothed Lasso propensity score model for large-scale covariate adjustment in healthcare database studies. _American Journal of Epidemiology_. 2024. Online ahead of print.



# Description

This package provides functions to fit undersmoothed Lasso models for propensity score weighting. The degree of undersmoothing is determined by balance criteria applied to the propensity score weighted cohorts. Balance critera for evaluating covariate balance include 

1) the average standardized absolute mean difference (ASAMD) 
2) the minimum of the largest standardized differences across all covariates.  

Options for propensity score weighting include 

1) inverse probability weighting 
2) overlap weighting 
3) matching weights

The package further provides diagnostics for evaluating model performance. In addition to balance and prediction diagnostics, we propose using synthetically generated negative control exposures for bias detection. The package provides a function for the process of generating synthetic negative control cohorts.  

# Overview of functions 

A brief description of each function in the package is provided below. A more detailed description for each function listed below is provided in separate vignettes.

* __treatment_model()__: 
  + calls glmnet to fit several Lasso models for treatment assignment with varying degrees of undersmoothing and returns a dataframe of the out-of-fold predictions for each fitted lasso model. 
* __balance_weighted_diff()__: 
  + calculates weighted standardized differences across treatment groups (called by *cov_diff_plot()*).
* __cov_diff_plot()__: 
  + plots standardized differences across treatment groups for all of the fitted lasso PS models after PS weighting 
* __ps_undersmooth_bal()__: 
  + takes as input a matrix of fitted PS values and selects the degree of undersmoothing using selected balance criteria.


## Installation

You can install the development version of PSLassoSynthNC from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("richiewyss/PSLassoSynthNC")
```

## Example

Below, we provide an example for running the full analytic pipeline from PS estimation, PS model selection/undersmoothing, treatment effect estimation, and diagnostic assessment for one simulation run. This pipeline was used in the publication by Wyss et al. (2025). 

Note: The code below is just for one simulation run. When running a simulation study (as in Wyss et al. (2025), the code needs to be run in a \'for\' loop (for i in 1:nsim) to get multiple simulation runs. To reduce computation time for simulation studies, each simluation run could be run in parallel with minor edits (e.g., using a parallel processing package). 

``` r
library(PSLassoSynthNC)
library(devtools)
devtools::load_all()
```

## Generate data

```r 
library(hal9001)
library(Matrix)
library(glmnet)
library(dplyr)

scenario<- 1
seed1<- 100       ## seed for simulating random data (when running in loop, should be different for each run)
seed2<- 110101    ## seed for setting global parameters for scenario 1 (should stay the same across runs)

dat<- dat_gen(n=5000, scenario=scenario, seed1=seed1, seed2=seed2)
```

## Running helper function to assign folds for cross-validation

``` r
nfolds = 10  
cvfolds <- stratifyCVFoldsByYandID(V=nfolds, Y = dat$A)
folds <- cvfolds$validRows
foldid <- cvfolds$fold_id
```

## Getting baseline covariates (for Scenario 2, requires fitting HAL on full data to extract design matrix)

``` r
x_names<- names(dat)[which(substr(names(dat), 1, 1)=='x')]
X_dat<- dat[,names(dat) %in% x_names]

if(scenario == 1){ x_basis_full<- X_dat }
if(scenario == 2){ x_basis_full<- hal_model(X=X_dat, Y=dat$A, max_degree=2, nfolds=NULL, foldid=foldid, num_knots=c(100, 25)) } 

## Reducing dimension of basis matrix (cleaning based on prevalence)
temp1<- apply(x_basis_full, 2, mean)
temp2<- (temp1 >= .01 & temp1 <= .99)
      
x_basis_full<- x_basis_full[,temp2]

## using treatment_model() function to fit glmnet on x_basis from above and extract predicted values for multiple lambda values

lasso_object<- treatment_model(data=x_basis_full, 
                               treatment=dat$A,
                               foldid=foldid,
                               alpha=1,
                               lambda_ratio=0.01, #ifelse(nrow(x_basis_full) < ncol(x_basis_full), 0.01, 1e-04)
                               nlambda=200,
                               nmodels=NULL,
                               maxit=5000,
                               penalty=NULL,
                               par=FALSE)

preds<- NULL
preds_cv<- NULL
preds_all<- NULL
coef_mat<- NULL
lambdas<- NULL
nvars<- NULL
      
preds<- lasso_object[[1]]       #same-sample predictions
preds_cv<- lasso_object[[2]]    #cross-validated (out-of-fold) predictions
coef_mat<- lasso_object[[3]]    #coefficient matrix (coefficients for each lambda)
lambdas<- lasso_object[[4]]     #lambda values
nvars<- lasso_object[[5]]       #number of variables selected by each lambda

##index values for subset of predictions and lambdas (select a range to reduce computation time instead of selecting all values)

steps<- NULL
steps<- floor(quantile(1:ncol(preds), seq(0, 1, .02))) 
    
preds_all<- preds_cv[,steps]
coef_mat<- coef_mat[,steps]
lambdas<- lambdas[steps]
nvars<- nvars[steps]
```


The function __main_fun()__ takes as input:

* _outcome_: 
  + a numeric vector representing the outcome.
* _treatment_: 
  + a vector of binary indicators representing treatment.
* _Xcovs_: 
  + a dataframe or matrix of baseline covariates.
* _model_type_: 
  + the type of lasso PS model. Options include 'lasso' or 'oal' for outcome adaptive lasso.
* _undersmooth_:
  + the method for selecting the degree of undersmoothing. Options include 'CV', 'balance', 'ctmle', or 'score'.
* _nmodels_
  + the number of lasso models to fit and consider for degrees of undersmoothing (must be between 1 and 12).
* _cross_fitting_
  + whether or not to use cross-fitted PS values (default is TRUE).
* _nfolds_
  + number of CV folds to use when fitting lasso models (default is 10).
* _trunc_type_
  + optional to truncate PS values or PS weights (default is NULL). Options include 'symmetric', 'assymetric', 'psvalue', 'psweight'.
* _trunc_val_low_
  + if 'trunc_type' is not null, then need to specify the lower value on which to truncate PS (between 0 and 1) 
* _trunc_val_high_
  + if 'trunc_type' is not null, then need to specify the upper value on which to truncate PS (between 0 and 1)
* _method_
  + the causal inference method used for estimating treatment effects. Options include 'ow', 'mw', 'ipw', 'smrw', 'matching', or 'tmle'.
  
```{r cars}
summary(cars)
```

You'll still need to render `README.Rmd` regularly, to keep `README.md` up-to-date. `devtools::build_readme()` is handy for this.

You can also embed plots, for example:

```{r pressure, echo = FALSE}
plot(pressure)
```

In that case, don't forget to commit and push the resulting figure files, so they display on GitHub and CRAN.
